Efficient simulation of individual-based population models

### 2024-04-02 article.qmd
update

sim_out_full <- popsim(model = model_full,
                       initial_population = pop_init,
                       events_bounds =c("birth" = 4 * params_birth$alpha,
                                        "death" = params_death$beta),
                       parameters = c(params_birth, params_death),
                       age_max = 2,
                       time = 500)

```{r echo=FALSE}

print(paste("The full algorithm is " ,format(sim_out_full$logs["duration_ns"]/sim_out$logs["duration_ns"], digits=2), " times slower than the randomized version", sep=""))

In @fig-fullvsrandom, the two algorithms are compared for different population sizes.  
We progressively decrease the value of the mortality rate parameter $\beta$ and increase the birth
rate parameter $\alpha$. Starting with the values provided
in @FerTra09, $\alpha=1$ and $\beta=2/300$, resulting in a stationary
population size of approximately $N=360$ individuals for a sample of 50
simulations, we can easily increase the stationary population size to
approximately $N=2600$ individuals with $\alpha=2$ and $\beta=1/300$
.^[The choices
    $(\alpha, \beta) \in \{(1,2/300),(1, 1/300), (1.5, 1/300), (2, 1/300)\}$
    lead to the stationary population sizes
    $N \in \{360, 900, 1800, 2600\}$. For each set of parameters, we
    generated a new initial population, which was used for a benchmark
    of 50 simulations with both randomized and full algorithm. The
    simulations run on a Intel Core i7-8550U CPU 1.80GHz × 8 processor,
    with 15.3 GiB of RAM, under Debian GNU/Linux 11.] In the log-scaled figure, we can observe the trend of computation
time as a function of the population size $N$, which is linear for the
randomized algorithm and quadratic for the full one (
 @alg-interaction). We can also see that the
randomized version of the algorithm is between 17 to 100 times faster
than the full one in this example, taking only 2 seconds in average for
the randomized version versus 211 seconds for 
@alg-interaction for the biggest population size
($N=2600$) and $T=500$.

>>>>

sim_out_full <- popsim(model = model_full,
    initial_population = pop_init,
    events_bounds =c("birth" = 4 * params_birth$alpha, "death" = params_death$beta),
    parameters = c(params_birth, params_death),
    age_max = 2,
    time = 500)
```

```{r echo=FALSE}
print(paste("The full algorithm is " ,format(sim_out_full$logs["duration_ns"]/sim_out$logs["duration_ns"], digits=2), " times slower than the randomized version", sep=""))
```

In @fig-fullvsrandom, the two algorithms are compared for different population sizes.  We progressively decrease the value of the mortality rate parameter $\beta$ and increase the birth rate parameter $\alpha$. Starting with the values provided in @FerTra09, $\alpha=1$ and $\beta=2/300$, resulting in a stationary population size of approximately $N=360$ individuals for a sample of 50 simulations, we can easily increase the stationary population size to approximately $N=2600$ individuals with $\alpha=2$ and $\beta=1/300$.^[The choices $(\alpha, \beta) \in \{(1,2/300),(1, 1/300), (1.5, 1/300), (2, 1/300)\}$ lead to the stationary population sizes $N \in \{360, 900, 1800, 2600\}$. For each set of parameters, we generated a new initial population, which was used for a benchmark of 50 simulations with both randomized and full algorithm. The simulations run on a Intel Core i7-8550U CPU 1.80GHz × 8 processor, with 15.3 GiB of RAM, under Debian GNU/Linux 11.] 
In the log-scaled figure, we can observe the trend of computation time as a function of the population size $N$, which is linear for the randomized algorithm and quadratic for the full one ( @alg-interaction). We can also see that the randomized version of the algorithm is between 17 to 100 times faster than the full one in this example, taking only 2 seconds in average for the randomized version versus 211 seconds for @alg-interaction for the biggest population size ($N=2600$) and $T=500$.
